---
title: "MB5370 Marine Genomics - Workshop 4"
subtitle: "Black Band Disease"
author: "Katie Dyck"
date: 2026-07-15
format:
  html:
      css: styles.css
      embed-resources: true
      code-fold: show
      code-copy: hover
      code-block-bg: true
      code-block-border-left: "grey25"
      toc: true
      number-sections: true
      number-depth: 4
      toc-location: left
      toc-expand: 3
      toc-title: Contents
      syntax-highlighting: github
      theme: cosmo
      title-block-banner: true
      fig-height: 6
      fig-width: 10
      fig-cap-location: bottom
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Background

### What is Black Band Disease? Why should we care about it?

Black Band Disease (BBD) is an infectious polymicrobial disease that impacts several coral species. It is identifiable by a dark band (black or dark red in colour) 1mm to several cm wide that surrounds a pale-coloured area where coral tissue has been lost. Depending on environmental conditions, the disease can spread from 1mm to 1cm per day until an entire coral colony is affected. The disease was discovered by Alfred Antonius in 1973 in Belize. Since then, the disease has been reported in shallow coral reefs throughout the Caribbean, the Indo-Pacific, the Red Sea, and the Great Barrier Reef. The fact that BBD continues to become more (rather than less) prevalent is cause for concern. Researchers have linked its increased prevalence to human activity and rising global temperatures. Reef-building corals - those that are of greatest importance to maintaining reef health - are most impacted by BBD.

### What do we know about the organisms that make up Black Band Disease and the taxonomic groups represented?

While researchers originally though it was caused by a single pathogen, it's now understood that BBD is is actually a complex microbial consortium that creates an anoxic, highly toxic (sulfur-rich) microenvironment coral tissue cannot survive in. There are 4 main taxonomic groups present in the consortium: *cyanobacteria*, *sulfate-reducing bacteria*, *sulfur-oxidising bacteria*, and *heterotrophs*.

1.  **Cyanobacteria** is the most prevalent taxa in the consortium. It's what causes the distinct dark-coloured band around lost tissue. They have multiple functions, including photosynthesis, carbon fixation, stabilisation of the microbial mat through production of extracellular polymeric substances (EPS), and oxygen production during the daytime. At night, when respiration exceeds photosynthesis, the mat becomes anoxic and tissue death occurs. While there are multiple strains of cyanobacteria associated with BBD, the most common is *Roseofilum reptotaenium*.
2.  **Sulfate-reducing bacteria**, present in the lower, oxygen-depleted regions of the microbial mat, are responsible for consuming oxygen, reducing sulfates, and generating hydrogen sulfide (one of the agents responsible for the death of coral tissue). Any tissue immediately below the dark band experience hypoxia, sulfide poisioning, and, eventually, death. $$\text{organic matter} + {SO}_4^{2-} \text{ (sulfate)} \rightarrow {CO}_2 + H_2S \text{ (hydrogen sulfide)}$$
3.  **Sulfur-oxidising bacteria** is present where there is both oxygen and sulfide. The bacteria oxidises hydrogen sulfide back into sulfur compounds: $$H_2S \rightarrow S^O \text{ (elemental sulfur)} \rightarrow {SO}_4^{2-}$$ Sulfides are not completely eliminated: substrates used by sulfate reducers are continuously regenerated, keeping sulfur cycling (and the death of affected coral tissue) going.
4.  **Heterotrophs** are responsible for the degradation of coral mucus (important for protection and removal of sediments), breakdown of coral tissue cell membranes, and dissolved organic matter. All of these activities contribute to nutrient recycling, the release of carbon substrates, and the production of electrons used by sulfate reducers.

Other taxonomic groups present include archaea, fungi, and other eukaryotic microbes.

There are distinct vertical environmental gradients within a BBD microbial mat.

-   At the *surface* is a large amount of light, oxygen (during the daytime), and cyanobacteria.
-   In the *middle* is the transition zone, which has sulfur-oxidising bacteria.
-   The *bottom* is an anoxic, sulfide-rich zone dominated by sulfate reducers. At night, photosynthesis stops, oxygen quickly disappears, and sulfide concentrations quickly increase.

### What are the gaps in our knowledge of BBD that could be addressed through an improved meta genome assembly?



### About the data

The data used in this workshop is a subset from fresh BBD samples from infected *Acropora hyacynthus* individuals at Orpheus Island. The molecular weight DNA was extracted and sequenced with an oxford nanopore MinION sequencer.

From this subset of data we are going to try to generate a metagenome assembly. We will analyse the BBD metagenome.

## Processing Log

### Step 1: Assembly

All outputs from this step will be saved inside a directory called `flye`, located in the main project directory.

```{bash flye directory, eval=FALSE, echo=FALSE}
mkdir flye
cd flye
```

Make a symbolic link to the BBD read data by entering the following line into the R terminal:

```{bash, eval=FALSE}
# make a symbolic link to BBD read data
ln -s /pvol/data/metagenome_assembly/bbd_0.1.fastq.gz .
```

Create a new shell script, `run_flye_bbd.sh`, inside the `flye` directory with the appropriate code. Submit the job for this script.

```{bash run_flye_bbd.sh code, eval=FALSE, echo=FALSE}
#!/bin/bash
#SBATCH --time=720
#SBATCH --ntasks=12 --mem=10gb

echo "Starting flye bbd in $(pwd) at $(date)"

flye --nano-hq ../raw_data/bbd_0.1.fastq.gz --threads 12 --out-dir bbd --meta

echo "Finished flye bbd in $(pwd) at $(date)"
```

```{bash submit job, eval=FALSE}
sbatch run_flye_bbd.sh

# `slurm-448.out` is my output file, saved in the flye directory.
```

### Step 2: Binning with metabat

1.  All outputs from this step will be saved inside a directory called `metabat`, located in the main project directory.
2.  Extract depth information.
3.  Reorder the contigs based on depth.
4.  Run metabat.

```{bash binning, eval=FALSE}
# Extract depth information
cat ../flye/bbd/assembly_info.txt  | grep -v '#' | awk '{OFS="\t";print $1,$3}' > depths_bbd.txt

# Reorder contigs based on depth
cat depths_bbd.txt | awk '{print $1}' | xargs samtools faidx ../flye/bbd/assembly.fasta > assembly_bbd_ordered.fasta

# Run metabat
metabat2 -i assembly_bbd_ordered.fasta -a depths_bbd.txt --cvExt -o bins_bbd/metabat --seed 5
```

The output from running metabat was: [**67 bins (39299051 bases in total) formed**]{style="font-family: 'Courier New', Courier, monospace;"}

### Step 3: Check bin quality with checkm

All outputs from this step will be saved inside a directory called `checkm`, located in the main project directory.

We want to find the number of the largest contig in the BBD assembly. We have to add this contig manually because it usually isn't included in the bins from metabat.

```{bash, eval=FALSE, echo=FALSE}
# Display lengths of all contigs in ascending order
cat ../flye/bbd/assembly.fasta | bioawk -c fastx '{print length($seq),$name}' | sort -n
```

-   The largest contig in our assembly is #119 (4,706,701 bp).

Within the `checkm` folder, create a new directory called `bins_bbd`.

```{bash, eval=FALSE}
cp ../metabat/bins_bbd/*.fa bins_bbd/
samtools faidx ../flye/bbd/assembly.fasta contig_119 > bins_bbd/contig_119.fa
```

Create a new slurm script, `run_checkm_bbd.sh`, and save it in the `checkm` directory. Submit the job through the terminal.

```{bash run_checkm_bbd slurm script, eval=FALSE, echo=FALSE}
#!/bin/bash
#SBATCH --time=60
#SBATCH --ntasks=6 --mem=40gb

echo "Starting checkm in $(pwd) at $(date)"

shopt -s expand_aliases
alias checkm='apptainer run -B /pvol/:/pvol /pvol/data/sif/checkm.sif checkm'

checkm lineage_wf bins_bbd out -t 4 -x fa -f checkm_results_bbd.txt

echo "Finished checkm in $(pwd) at $(date)"
```

### Step 4: Select bins based on checkm results

The results of the checkm bin quality check can be found in the text file `checkm_results_bbd.txt`, which is saved inside the checkm directory.

We want to look at the most complete bins as they are most likely to contain most of the useful information on the function of BBD. These bins tend to correspond to the most abundant (and best assembled) members of the community. We will use the "Completeness" column in `checkm_results_bbd.txt` to assess these results. We will focus on bins with a completeness greater than 15%.

```{bash, eval=FALSE}
cat checkm_results_bbd.txt | awk '$13>15'

# We are only concerned with the bin ID column
cat checkm_results_bbd.txt | awk '$13>15{print $1}'
```

Save the output of the terminal commands above to a file, `bbd_topbins.txt`, so we can continue using it in our analysis:

```{bash, eval=FALSE}
cat checkm_results_bbd.txt | awk '$13>15{print $1}' > bbd_topbins.txt
```

### Step 5: Perform taxonomic classification with gtdbtk

All outputs from this step will be saved inside a directory called `gtdbtk`, located in the main project directory.

-   We'll create a subfolder within the `gtdbtk` folder called `genomes`. Here we will save a copy of each of the top bins identified in step 4.

```{bash, eval=FALSE}
# Iterate over our list of top bins
# For each bin we want to copy a file corresponding to that bin into the `genomes` folder
while read bin;do echo $bin;done < ../checkm/bbd_topbins.txt
while read bin;do echo "../metabat/bins_bbd/${bin}.fa";done < ../checkm/bbd_topbins.txt
while read bin;do cp "../metabat/bins_bbd/${bin}.fa" genomes;done < ../checkm/bbd_topbins.txt
```

Now we will run the classification workflow. Create a shell script inside the `gtdbtk` folder called `run_gtdbtk.sh` and submit it through the terminal.

```{bash run_gtdbtk shell script, eval=FALSE, echo=FALSE}
#!/bin/bash
#SBATCH --time=60
#SBATCH --ntasks=2 --mem=40gb

echo "Starting gtdbtk in $(pwd) at $(date)"

shopt -s expand_aliases
alias gtdbtk='apptainer run -B /pvol:/pvol,/pvol/data/mg/gtdbtkdata/release207_v2/:/refdata /pvol/data/sif/gtdbtk.sif'

gtdbtk classify_wf -x fa --cpus 2 --genome_dir genomes --out_dir out --skip_ani_screen

echo "Finished gtdbtk in $(pwd) at $(date)"
```

The output of this file is saved in `gtdbtk.bac120.summary.tsv`.

### Step 6: Find and annotate genes with prokka

For us to better our understanding of metabolic processes within the BBD metagenome, we need to identify which genes are present within the genomes of the BBD taxa. The program [Prokka](https://github.com/tseemann/prokka) will identify gene sequences and homologous genes with known functions for us. Homologous genes share a common ancestor, therefore their functions are probably similar.

All outputs from this step will be saved inside a directory called `prokka`, located in the main project directory.

First, we want to run prokka for a single bin. Identify the first bin ID from `bbd_topbins.txt`, then run the following code through the terminal:

```{bash, eval=FALSE}
bin=contig_119
prokka --outdir $bin --prefix $bin  ../metabat/bins_bbd/${bin}.fa
```

Running this creates a folder `contig_119`, which contains the output file `contig_119.tsv` with the name of each gene and its name and functional information.

Create a blank shell script called `run_prokka.sh` in the `prokka` directory with the appropriate code. Run the job through the terminal.

```{bash run_prokka shell script, eval=FALSE, echo=FALSE}
#!/bin/bash
#SBATCH --time=60
#SBATCH --ntasks=2 --mem=2gb

echo "Starting prokka in $(pwd) at $(date)"

shopt -s expand_aliases
alias prokka='apptainer run -B /pvol/:/pvol /pvol/data/sif/prokka.sif prokka'

while read bin;do 
  prokka --outdir $bin --prefix $bin "../metabat/bins_bbd/${bin}.fa";
done < ../checkm/bbd_topbins.txt

echo "Finished prokka in $(pwd) at $(date)"
```

```{bash, eval=FALSE}
# Output file is named "slurm-603.out"
```

### Step 8: Infer metabolic pathways

To take the functional information for individual genes in contig_119 and use it to infer the presence of broader biological pathways (i.e. multiple genes), we will use the tool [minpath](https://omics.informatics.indiana.edu/MinPath/). We will run this on each genome in our analysis.

All outputs from this step will be saved inside a directory called `minpath`, located in the main project directory.

We will use the `contig_119.tsv` file to prepare inputs for minpath. Within this file:

-   *EC_number* is Enzyme Commission number (e.g. 2.5.1.41)
-   *COG* is Clustes of Orthologous Genes (e.g. COG1646)
-   *product* is the protein name (e.g. Geranylgeranylglyceryl phosphate synthase)

We are only interested in lines with an EC number. This output is saved in the file `contig_119.ec`. We run minpath on this file by entering this command in the terminal:

```{bash, eval=FALSE}
minpath -ec contig_119.ec -report contig_119.ec.report -details contig_119.ec.details
```

In the `contig_119.ec.report`:

-   *PWY-7321* is a pathway identifier in the metacyc database. It can be used to look up the full pathway (E.g. [PWY-7321](https://biocyc.org/pathway?orgid=META&id=PWY-7321))
-   *fam0 4* tells us four gene families are involved in ths=is pathway
-   *fam-found 2* tells us we found two gene families from the pathway in this genome
-   *name ecdysteroid metabolism (arthropods)* is the pathway name

Run the following code in the terminal (once the prokka job is finished running) to prepare minpath inputs for each bin:

```{bash prep minpath inputs, eval=FALSE}
while read bin;do 
  cat ../prokka/$bin/${bin}.tsv | awk -F '\t' '$5{printf("%s\t%s\n", $1,$5)}' | grep -v '^locus' > ${bin}.ec 
done < ../checkm/bbd_topbins.txt
```

Next, run the code below in the terminal to run minpath on every bin:

```{bash run minpath on bins, eval=FALSE}
while read bin;do 
  minpath -ec ${bin}.ec -report ${bin}.ec.report -details ${bin}.ec.details
done < ../checkm/bbd_topbins.txt
```

Two pathways from the output are of interest to us.

1.  **Photosynthesis light reactions** (metacyc code `PWY-101`) - this includes all the components necessary for photosynthesis that take place in light, Photosystem I, and Photosystem II.

```{bash search photosynthesis, eval=FALSE}
grep 'PWY-101' *.ec.report
```

`PWY-101` is found in contig_119, metabat.27, metabat.48, and metabat.50.

To view the taxonomy of bins containing this pathway, cross reference the results with the output from `gtdbtk`.

```{bash view bin taxonomy, eval=FALSE}
grep 'contig_119' ../gtdbtk/out/gtdbtk.bac120.summary.tsv
grep 'metabat.27' ../gtdbtk/out/gtdbtk.bac120.summary.tsv
grep 'metabat.48' ../gtdbtk/out/gtdbtk.bac120.summary.tsv
grep 'metabat.50' ../gtdbtk/out/gtdbtk.bac120.summary.tsv
```

2.  [**Superpathway of tetrathionate reduction**](https://biocyc.org/META/NEW-IMAGE?type=PATHWAY&object=PWY-5360) (metacyc code `PWY-5360`) - includes enzymes needed to reduce sulfite to hydrogen sulfide.

```{bash search tetrathionate, eval=FALSE}
grep 'PWY-5360' *.ec.report
```

`PWY-5360` is only found in metabat.56.

We will cross-reference the results with the `gtdbtk` output to view the taxonomy of bins containing this pathway.

```{bash view tetrathionate bin taxonomy, eval=FALSE}
# Cross reference results with gtdbtk output
grep 'metabat.56' ../gtdbtk/out/gtdbtk.bac120.summary.tsv
```

## Results

### Summarise your metagenome assembly of black band disease. How good is the assembly? Think about the metrics you have learned throughout the module to report on this.

In my metagenome assembly, the most complete bins are *contig_119* and *metabat.50* (97.62% and 97.56%, respectively). In the contig_119 bin, only 9 of 454 markers are missing, while in the metabat.50 bin only 11 of the 579 markers are missing. The high completeness of these bins indicates that they likely contain most of the information pertaining to function within the BBD community.

### What organisms make up the largest components of the assembly? (results from gtdbtk)

I have joined the appropriate columns from `checkm_results_bbd.txt` with `gtdbtk.bac120.summary.tsv` to help visualise which organisms make up the largest components of the assembly.

```{r load packages, echo=FALSE, include=FALSE}
library(tidyverse)
library(readr)
```

```{r gtdbtk results, include=FALSE}
gtdbtk_result <- read_tsv("../gtdbtk/out/gtdbtk.bac120.summary.tsv")
# head(gtdbtk_result)

checkm_result <- readLines("../checkm/checkm_results_bbd.txt")
lines <- checkm_result[!grepl("^-+$", checkm_result)]   # Remove dashed lines
checkm_result <- read_table(paste(lines, collapse = "\n"))   # Read using whitespace as the delimiter
# head(checkm_result)
```

```{r left join}
# Left join
org_assembly <- left_join(gtdbtk_result, checkm_result, by = c("user_genome" = "Bin")) %>%
  # remove unnecessary columns
  select(c(1,2,23,24,26:33)) %>%
  # fix column names
  rename(c("bin_id" = "user_genome", "organism" = "classification", "# genomes" = "lineage", "# markers" = "#", "0" = "#_1", "1" = "markers", "2" = "#_2", "3" = "marker", "4" = "sets", "5+" = "0", "completeness" = "1", "contamination" = "2"))

print(org_assembly, n = Inf)
```

### A brief summary of the results of prokka and minpath annotation on this assembly. How many genes were identified by Prokka? Of these how many were assigned an EC number?

```{bash, eval=FALSE}
cd prokka
ls
ls contig_119
ls contig_119/contig_119.tsv

# while loop reads each line, assigns variable bin to be the value on each line
while read bin;do echo $bin; done < ../checkm/bbd_topbins.txt
# each one is a directory in our prokka directory. we want a pattern line (e.g. name_of_bin, name_of_bin.tsv).

cat ../minpath/*.ec|wc -l
```

## Discussion

**Comment on the quality of your assembly compared with previously published results using Illumina data from Sato et al. Can you identify bins representing key bacterial taxa that are found in both assemblies? Are there any highly complete bins in one assembly that do not have a clear match in the other assembly? For bins that match across assemblies how to the completeness and contamination statistics compare?**

**Comment on the results of pathway analysis with minpath. Which bin contains most of the genes required for photosynthesis? Which bin or bins are most likely to be contributing to sulfate reduction?**
