---
title: "Module 3: Marine Genomics"
subtitle: "Black Band Disease"
author: "Katie Dyck"
date: 2026-07-15
format:
  html:
      df-scroll: true
      df-print: paged
      css: styles.css
      embed-resources: true
      code-fold: show
      code-copy: hover
      code-block-bg: true
      code-block-border-left: "grey25"
      toc: true
      number-sections: true
      number-depth: 4
      toc-location: left
      toc-expand: 3
      toc-title: Contents
      syntax-highlighting: github
      theme: cosmo
      title-block-banner: true
      fig-height: 6
      fig-width: 10
      fig-cap-location: bottom
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Background

**What is Black Band Disease? Why should we care about it?**

Black Band Disease (BBD) is an infectious polymicrobial disease that impacts several reef-building coral species. It is identifiable by a distinct dark band (black or dark red in colour) ranging from 1mm to several cm wide, which surrounds the pale, exposed skeleton where coral tissue has died. The disease spreads as the microbial band moves across the coral colony, spreading at a rate of 1mm to 1cm per day depending on environmental conditions. Researchers studying BBD have confirmed that higher temperatures cause the disease to spread at a faster rate, and become more virulent. Should the disease progress enough, it can kill an entire coral colony.

Since it was first discovered by Alfred Antonius in Belize in 1973, BBD has been reported in coral reefs throughout the Caribbean, Indo-Pacific, Red Sea, and Great Barrier Reef. The fact that BBD continues to become more (rather than less) prevalent is cause for concern. Research has shown that the disease is caused by a variety of microorganisms, including cyanobacteria, sulfate-reducing bacteria, sulfur-oxidising bacteria, and heterotrophs, rather than a single pathogen. Seasonal outbreaks are tied to increased water temperatures and high light levels; the disease becomes more prevalent and progresses faster in warmer summer months. Studies have shown that higher temperatures can also increase the rate of the disease progression and virulence.

BBD is worth caring about because it adds to the growing list of factors impacting coral reef health worldwide, including climate change, pollution, ocean acidification, and human impact. Reef-building corals provide the structural foundation of reef ecosystems, supporting a diversity of marine species. As global temperatures continue to increase, BBD outbreaks will only become more frequent and severe, causing further degradation to already fragile reef ecosystems. It is crucial that we understand the biology, environmental drivers and microbial interactions of BBD so we can predict outbreaks and create strategies to protect coral reefs worldwide.

**What do we know about the organisms that make up Black Band Disease and the taxonomic groups represented?**

While researchers originally though it was caused by a single pathogen, it's now understood that BBD is is actually a complex microbial consortium that creates an anoxic, highly toxic (sulfur-rich) microenvironment coral tissue cannot survive in. There are 4 main taxonomic groups present in the consortium: *cyanobacteria*, *sulfate-reducing bacteria*, *sulfur-oxidising bacteria*, and *heterotrophs*.

1.  **Cyanobacteria** is the most prevalent taxa in the consortium. It's what causes the distinct dark-coloured band around lost tissue. They have multiple functions, including photosynthesis, carbon fixation, stabilisation of the microbial mat through production of extracellular polymeric substances (EPS), and oxygen production during the daytime. At night, when respiration exceeds photosynthesis, the mat becomes anoxic and tissue death occurs. While there are multiple strains of cyanobacteria associated with BBD, the most common is *Roseofilum reptotaenium*.
2.  **Sulfate-reducing bacteria**, present in the lower, oxygen-depleted regions of the microbial mat, are responsible for consuming oxygen, reducing sulfates, and generating hydrogen sulfide (one of the agents responsible for the death of coral tissue). Any tissue immediately below the dark band experience hypoxia, sulfide poisioning, and, eventually, death. $$\text{organic matter} + {SO}_4^{2-} \text{ (sulfate)} \rightarrow {CO}_2 + H_2S \text{ (hydrogen sulfide)}$$
3.  **Sulfur-oxidising bacteria** is present where there is both oxygen and sulfide. The bacteria oxidises hydrogen sulfide back into sulfur compounds: $$H_2S \rightarrow S^O \text{ (elemental sulfur)} \rightarrow {SO}_4^{2-}$$ Sulfides are not completely eliminated: substrates used by sulfate reducers are continuously regenerated, keeping sulfur cycling (and the death of affected coral tissue) going.
4.  **Heterotrophs** are responsible for the degradation of coral mucus (important for protection and removal of sediments), breakdown of coral tissue cell membranes, and dissolved organic matter. All of these activities contribute to nutrient recycling, the release of carbon substrates, and the production of electrons used by sulfate reducers.

Other taxonomic groups present include archaea, fungi, and other eukaryotic microbes.

There are distinct vertical environmental gradients within a BBD microbial mat.

-   At the *surface* is a large amount of light, oxygen (during the daytime), and cyanobacteria.
-   In the *middle* is the transition zone, which has sulfur-oxidising bacteria.
-   The *bottom* is an anoxic, sulfide-rich zone dominated by sulfate reducers. At night, photosynthesis stops, oxygen quickly disappears, and sulfide concentrations quickly increase.

**What are the gaps in our knowledge of BBD that could be addressed through an improved meta genome assembly?**

There are several gaps in our understanding of BBD that an improved metagenome assembly could help address. Most BBD studies rely on short-read genomic sequencing and taxonomic classification of individual reads. While these approaches have successfully identified which microorganisms are associated with BBD, they usually cannot reliably link metabolic genes to specific organisms because genomes stay fragemented. Improved metagenome assembly would generate more complete metagenome-assembled genomes (MAGs), allowing researchers to determine which organisms (e.g. cyanobacteria, heterotrophs) are responsible for which metabolic processes such as sulfur reduction, toxin production, virulence, and nutrient cycling. Current studies can tell us which organisms are present, but improved assemblies would help reveal what each organism is capable of doing.

Although *Roseofilum reptotaenium* has consistently been associated as the dominant cyanobacteria in BBD mats, there remains uncertainty around its role in the progression of the disease. For example, it's still unclear whether *R. reptotaenium* damages coral tissue directly or creates an environment that allows other microorganisms to become pathogenic. In addition, genetic and metabolic diversity among *Roseofilum* strains from different geographic regions is poorly understood. More complete metagenome assemblies could reconstruct high-quality *Roseofilum* genomes, allowing researchers to compare strains from different regions and identify genes responsibly for pathogen production, toxin production, and environmental adaptation.

Our understanding of sulfur cycling in BBD mats is largely incomplete. Sulfate-reducing bacteria produces hydrogen sulfide, while sulfur-oxidising bacteria reduces sulfur compounds, but complete metabolic pathways and organisms responsible for each step in the pathway have not been completely identified. Higher-quality metagenome assemblies would allow reconstruction of complete sulfur metabolism pathways and help scientists identify how interactions between sulfur-oxidising and sulfate-reducing microorganisms maintain the sulfide-rich environment that causes coral tissue death.

Many metagenome reads from BBD-infected corals are unable to be assigned to known species because reference genomes are incomplete. Improved metagenome assemblies could recover genomes from previously unknown microorganisms, improve current BBD databases, and provide scientists with a better understanding of microbial diversity in BBD communities around the world.

### About the data

The data used in this workshop is a subset from a JCU student's PhD project, which was published in 2025 (the paper can be found [here](https://doi.org/10.1111/1462-2920.70219)). They obtained fresh BBD samples from infected *Acropora hyacynthus* individuals at Orpheus Island. The molecular weight DNA was extracted and sequenced with an oxford nanopore MinION sequencer. We will be working with a subset of the complete data set with the intention of analysing the black band disease metagenome. 

## Processing Log

::: {.callout-important icon=FALSE}
In this document, all yellow code chunks indicate bash (R Terminal) code input. Comments begin with `#`; these are listed for reference only, and should not be entered into the Terminal.

All other code chunks indicate R code, which should be input into the Console.
:::

### Step 1: Assembly

`flye`'s job is to look at each DNA sequence and assemble them into longer sequences called a **contig**.

(All outputs from this step will be saved inside a directory called `flye`, located in the main project directory.)

```{bash flye directory, eval=FALSE, echo=FALSE}
mkdir flye
cd flye
```

Make a symbolic link (shortcut) to the BBD raw sequencing reads data (`bbd_0.1.fastq.gz .`) by entering the following line into the R terminal:

```{bash, eval=FALSE}
#| code-fold: false

ln -s /pvol/data/metagenome_assembly/bbd_0.1.fastq.gz .
```

Create a new shell script, [`run_flye_bbd.sh`](https://github.com/katiedyck/Marine-Genomics/blob/main/flye/run_flye_bbd.sh){target="_blank"}, inside the `flye` directory with the appropriate code. Submit the job for this script.

```{bash run_flye_bbd.sh code, eval=FALSE, echo=FALSE}
#!/bin/bash
#SBATCH --time=720
#SBATCH --ntasks=12 --mem=10gb

echo "Starting flye bbd in $(pwd) at $(date)"

flye --nano-hq ../raw_data/bbd_0.1.fastq.gz --threads 12 --out-dir bbd --meta

echo "Finished flye bbd in $(pwd) at $(date)"
```

```{bash submit job, eval=FALSE}
#| code-fold: false

sbatch run_flye_bbd.sh

# `slurm-448.out` is my output file
```

Before assembly, we just had millions of disconnected reads and information about which taxa are present. After assembly, we have much longer DNA sequences that can be grouped into MAGs, annotated to identify genes and metabolic pathways, and used to determine which organisms carry genes for sulfur metabolism, toxin production, or other functions. This is why metagenome assembly is a crucial step: it bridges the gap between identifying who is present in the BBD community and understanding what those organisms are capable of doing.

### Step 2: Binning with metabat

(All outputs from this step will be saved inside a directory called `metabat`, located in the main project directory.)

The goal of this step is to go from contigs to individual microbial genomes. The workflow can be visualised like this: $$\text{Raw DNA reads} \rightarrow \text{Assembly} \rightarrow \text{Contigs} \rightarrow \text{Binning} \rightarrow \text{Metagenome-Assembled Genomes (MAGs)}$$ Right now we have a file with thousands of contigs ([`assembly_info.txt`](https://github.com/katiedyck/Marine-Genomics/blob/main/flye/bbd/assembly_info.txt){target="_blank"}) and we don't know which organisms belong with which contig. We will use metabat to sort the contigs into "bins". The hope is that each bin represents one genome.

```{bash binning, eval=FALSE}
#| code-fold: false

# Extract the depth information for each contig
cat ../flye/bbd/assembly_info.txt  | grep -v '#' | awk '{OFS="\t";print $1,$3}' > depths_bbd.txt
```

-   **Depth** is the average number of sequencing reads that cover any given region in a contig. (For example, a contig that has been sequenced 10 times has a greater depth than a contig that has only been sequenced 2 times.) It is useful for helping metabat group contigs together as contigs with very similar depths are most likely from the same organism.

```{bash, eval=FALSE}
#| code-fold: false

# Reorder contigs based on their depth
cat depths_bbd.txt | awk '{print $1}' | xargs samtools faidx ../flye/bbd/assembly.fasta > assembly_bbd_ordered.fasta

# Run metabat
metabat2 -i assembly_bbd_ordered.fasta -a depths_bbd.txt --cvExt -o bins_bbd/metabat --seed 5
```

-   The output from running metabat was: [**67 bins (39299051 bases in total) formed**]{style="font-family: 'Courier New', Courier, monospace;"}
-   The depth of each contig is saved in [`depths_bbd.txt`](https://github.com/katiedyck/Marine-Genomics/blob/main/metabat/depths_bbd.txt){target="_blank"}.

### Step 3: Check bin quality with checkm

(All outputs from this step will be saved inside a directory called `checkm`, located in the main project directory.)

Now that we've created our bins, we want to check how good they are. We want to find the number of the largest contig in the BBD assembly. We have to add this contig manually because it usually isn't included in the bins from metabat due to its unusual sequence composition or repetitive DNA sequences.

```{bash, eval=FALSE, echo=FALSE}
# Display lengths of all contigs in ascending order
cat ../flye/bbd/assembly.fasta | bioawk -c fastx '{print length($seq),$name}' | sort -n
```

-   **The largest contig in our assembly is contig_119 (4,706,701 bp).**

Within the `checkm` folder, create a new directory called `bins_bbd`.

Run this code in the R terminal so that checkm will analyse contig_119 alongside the other metabat bins.

```{bash, eval=FALSE}
#| code-fold: false

# Copy the metabat bins into the 'bins_bbd' folder
cp ../metabat/bins_bbd/*.fa bins_bbd/

# Extract the largest contig in the assembly and save it as its own fasta file (.fa)
samtools faidx ../flye/bbd/assembly.fasta contig_119 > bins_bbd/contig_119.fa
```

Create a new slurm script, [`run_checkm_bbd.sh`](https://github.com/katiedyck/Marine-Genomics/blob/main/checkm/run_checkm_bbd.sh){target="_blank"}, and save it in the `checkm` directory. Submit the job through the terminal.

```{bash run_checkm_bbd slurm script, eval=FALSE, echo=FALSE}
#!/bin/bash
#SBATCH --time=60
#SBATCH --ntasks=6 --mem=40gb

echo "Starting checkm in $(pwd) at $(date)"

shopt -s expand_aliases
alias checkm='apptainer run -B /pvol/:/pvol /pvol/data/sif/checkm.sif checkm'

checkm lineage_wf bins_bbd out -t 4 -x fa -f checkm_results_bbd.txt

echo "Finished checkm in $(pwd) at $(date)"
```

### Step 4: Select bins based on checkm results

Now we want to choose which genomes are worth analysing in detail.

The results of the checkm bin quality check can be found in the text file [`checkm_results_bbd.txt`](https://github.com/katiedyck/Marine-Genomics/blob/main/checkm/checkm_results_bbd.txt){target="_blank"}, saved inside our checkm directory.

**A good genome has high completeness and low contamination.**

We want to look at the most complete bins as they are most likely to contain most of the useful information on the function of BBD. These bins tend to correspond to the most abundant (and best assembled) members of the community. We will use the "Completeness" column in `checkm_results_bbd.txt` to assess these results. We will focus on bins with a completeness greater than 15%. Run the code below through the R terminal to find these bins.

```{bash, eval=FALSE}
#| code-fold: false

cat checkm_results_bbd.txt | awk '$13>15'

# We are only concerned with the bin ID column
cat checkm_results_bbd.txt | awk '$13>15{print $1}'
```

Save the output of these commands to a file, [`bbd_topbins.txt`](https://github.com/katiedyck/Marine-Genomics/blob/main/checkm/bbd_topbins.txt){target="_blank"}, so we can continue using it in our analysis.

```{bash, eval=FALSE}
#| code-fold: false

cat checkm_results_bbd.txt | awk '$13>15{print $1}' > bbd_topbins.txt
```

### Step 5: Perform taxonomic classification with gtdbtk

All outputs from this step will be saved inside a directory called `gtdbtk`, located in the main project directory.

-   We'll create a subfolder within the `gtdbtk` folder called `genomes`. Here we will save a copy of each of the top bins identified in step 4.

```{bash, eval=FALSE}
#| code-fold: false

# Iterate over our list of top bins
# For each bin we want to copy a file corresponding to that bin into the `genomes` folder
while read bin;do echo $bin;done < ../checkm/bbd_topbins.txt
while read bin;do echo "../metabat/bins_bbd/${bin}.fa";done < ../checkm/bbd_topbins.txt
while read bin;do cp "../metabat/bins_bbd/${bin}.fa" genomes;done < ../checkm/bbd_topbins.txt
```

Now we will run the classification workflow. Create a shell script inside the `gtdbtk` folder called [`run_gtdbtk.sh`](https://github.com/katiedyck/Marine-Genomics/blob/main/gtdbtk/run_gtdbtk.sh){target="_blank"}, insert the appropriate code, and submit it through the terminal.

```{bash run_gtdbtk shell script, eval=FALSE, echo=FALSE}
#!/bin/bash
#SBATCH --time=60
#SBATCH --ntasks=2 --mem=40gb

echo "Starting gtdbtk in $(pwd) at $(date)"

shopt -s expand_aliases
alias gtdbtk='apptainer run -B /pvol:/pvol,/pvol/data/mg/gtdbtkdata/release207_v2/:/refdata /pvol/data/sif/gtdbtk.sif'

gtdbtk classify_wf -x fa --cpus 2 --genome_dir genomes --out_dir out --skip_ani_screen

echo "Finished gtdbtk in $(pwd) at $(date)"
```

The output of this file is saved in [`gtdbtk.bac120.summary.tsv`](https://github.com/katiedyck/Marine-Genomics/blob/main/gtdbtk/out/classify/gtdbtk.bac120.summary.tsv){target="_blank"}.

### Step 6: Find and annotate genes with prokka

For us to better our understanding of metabolic processes within the BBD metagenome, we need to identify which genes are present within the genomes of the BBD taxa. The program [Prokka](https://github.com/tseemann/prokka){target="_blank"} will identify gene sequences and homologous genes with known functions for us. Homologous genes share a common ancestor, therefore their functions are probably similar.

All outputs from this step will be saved inside a directory called `prokka`, located in the main project directory.

First, we want to run prokka for a single bin. Identify the first bin ID from `bbd_topbins.txt`, then run the following code through the terminal:

```{bash, eval=FALSE}
#| code-fold: false

bin=contig_119
prokka --outdir $bin --prefix $bin  ../metabat/bins_bbd/${bin}.fa
```

Running this creates a folder `contig_119`, which contains the output file `contig_119.tsv` with the name of each gene and its name and functional information.

Create a blank shell script called `run_prokka.sh` in the `prokka` directory with the appropriate code. Run the job through the terminal.

```{bash run_prokka shell script, eval=FALSE, echo=FALSE}
#!/bin/bash
#SBATCH --time=60
#SBATCH --ntasks=2 --mem=2gb

echo "Starting prokka in $(pwd) at $(date)"

shopt -s expand_aliases
alias prokka='apptainer run -B /pvol/:/pvol /pvol/data/sif/prokka.sif prokka'

while read bin;do 
  prokka --outdir $bin --prefix $bin "../metabat/bins_bbd/${bin}.fa";
done < ../checkm/bbd_topbins.txt

echo "Finished prokka in $(pwd) at $(date)"
```

```{bash, eval=FALSE, include=FALSE}
# Output file is named "slurm-603.out"
```

### Step 8: Infer metabolic pathways

To take the functional information for individual genes in *contig_119* and use it to infer the presence of broader biological pathways (i.e. multiple genes), we will use the tool [minpath](https://omics.informatics.indiana.edu/MinPath/){target="_blank"}. We will run this on each genome in our analysis.

All outputs from this step will be saved inside a directory called `minpath`, located in the main project directory.

We will use the `contig_119.tsv` file to prepare inputs for minpath. Within this file:

-   *EC_number* is Enzyme Commission number (e.g. 2.5.1.41)
-   *COG* is Clustes of Orthologous Genes (e.g. COG1646)
-   *product* is the protein name (e.g. Geranylgeranylglyceryl phosphate synthase)

We are only interested in lines with an EC number. This output is saved in the file `contig_119.ec`. We run minpath on this file by entering this command in the terminal:

```{bash, eval=FALSE}
#| code-fold: false

minpath -ec contig_119.ec -report contig_119.ec.report -details contig_119.ec.details
```

In the `contig_119.ec.report`:

-   *PWY-7321* is a pathway identifier in the metacyc database. It can be used to look up the full pathway (E.g. [PWY-7321](https://biocyc.org/pathway?orgid=META&id=PWY-7321){target="_blank"})
-   *fam0 4* tells us four gene families are involved in ths=is pathway
-   *fam-found 2* tells us we found two gene families from the pathway in this genome
-   *name ecdysteroid metabolism (arthropods)* is the pathway name

Run the following code in the terminal (once the prokka job is finished running) to prepare minpath inputs for each bin:

```{bash prep minpath inputs, eval=FALSE}
#| code-fold: false

while read bin;do 
  cat ../prokka/$bin/${bin}.tsv | awk -F '\t' '$5{printf("%s\t%s\n", $1,$5)}' | grep -v '^locus' > ${bin}.ec 
done < ../checkm/bbd_topbins.txt
```

Next, run the code below in the terminal to run minpath on every bin:

```{bash run minpath on bins, eval=FALSE}
#| code-fold: false

while read bin;do 
  minpath -ec ${bin}.ec -report ${bin}.ec.report -details ${bin}.ec.details
done < ../checkm/bbd_topbins.txt
```

Two pathways from the output are of interest to us.

1.  **Photosynthesis light reactions** (metacyc code `PWY-101`) - this includes all the components necessary for photosynthesis that take place in light, Photosystem I, and Photosystem II.

```{bash search photosynthesis, eval=FALSE}
#| code-fold: false

grep 'PWY-101' *.ec.report
```

`PWY-101` is found in contig_119, metabat.27, metabat.48, and metabat.50.

To view the taxonomy of bins containing this pathway, cross reference the results with the output from `gtdbtk`.

```{bash view bin taxonomy, eval=FALSE}
#| code-fold: false

grep 'contig_119' ../gtdbtk/out/gtdbtk.bac120.summary.tsv
grep 'metabat.27' ../gtdbtk/out/gtdbtk.bac120.summary.tsv
grep 'metabat.48' ../gtdbtk/out/gtdbtk.bac120.summary.tsv
grep 'metabat.50' ../gtdbtk/out/gtdbtk.bac120.summary.tsv
```

2.  [**Superpathway of tetrathionate reduction**](https://biocyc.org/META/NEW-IMAGE?type=PATHWAY&object=PWY-5360){target="_blank"} (metacyc code `PWY-5360`) - includes enzymes needed to reduce sulfite to hydrogen sulfide.

```{bash search tetrathionate, eval=FALSE}
#| code-fold: false

grep 'PWY-5360' *.ec.report
```

`PWY-5360` is only found in metabat.56.

We will cross-reference the results with the `gtdbtk` output to view the taxonomy of bins containing this pathway.

```{bash view tetrathionate bin taxonomy, eval=FALSE}
#| code-fold: false

grep 'metabat.56' ../gtdbtk/out/gtdbtk.bac120.summary.tsv
```

## Results

1. **Summarise your metagenome assembly of black band disease. How good is the assembly? Think about the metrics you have learned throughout the module to report on this.**

The metagenome assembly produced 67 bins comprising approximately 39.3 Mb of assembled sequence. The highest-quality genomes (i.e. most complete bins) were `contig_119` (97.62% completeness) and `metabat.50` (97.56% completeness), indicating that these genomes are nearly complete. High completeness suggests that most expected marker genes were recovered; these genomes contain majority of the information pertaining to function within the BBD community. According to CheckM, only 9 of 454 marker genes were missing from contig_119 and 11 of 579 marker genes were missing from metabat.50. Contamination levels were low (1.64% for contig_119 and 0.67% for metabat.50), these represent high-quality MAGs.

2. **What organisms make up the largest components of the assembly? (results from gtdbtk)**

I have joined the appropriate columns from `checkm_results_bbd.txt` with `gtdbtk.bac120.summary.tsv` to help visualise which organisms make up the largest components of the assembly.

```{r load packages, echo=FALSE, include=FALSE}
library(tidyverse)
library(readr)
library(DT)
```

```{r gtdbtk results, echo=FALSE, include=FALSE}
gtdbtk_result <- read_tsv("../gtdbtk/out/gtdbtk.bac120.summary.tsv")
# head(gtdbtk_result)

checkm_result <- readLines("../checkm/checkm_results_bbd.txt")
lines <- checkm_result[!grepl("^-+$", checkm_result)]   # Remove dashed lines
checkm_result <- read_table(paste(lines, collapse = "\n"))   # Read using whitespace as the delimiter
# head(checkm_result)
```

```{r left join}
#| code-fold: true
# Left join
org_assembly <- left_join(gtdbtk_result, checkm_result, by = c("user_genome" = "Bin")) %>%
  # remove unnecessary columns
  select(c(1,2,23,24,26:33)) %>%
  # fix column names
  rename(c("Bin ID" = "user_genome", "Organism" = "classification", "# Genomes" = "lineage", "# Markers" = "#", "0" = "#_1", "1" = "markers", "2" = "#_2", "3" = "marker", "4" = "sets", "5+" = "0", "Completeness %" = "1", "Contamination %" = "2"))

datatable(org_assembly, options = list(scrollX = TRUE, scrollCollapse = TRUE, paging = FALSE), rownames = FALSE,
          caption = "GTDB-Tk and CheckM results for the top 16 genomes in the assembly.")
```

<br/> On all 16 bins that are above 15% completeness, they cluster into 5 phyla, with sulfur-cycling bacteria dominating by bin count:

-   **Desulfobacterota (5 bins)** — Sulfate/sulfite reducers. Present in metabat.27, metabat.23, metabat.41, metabat.42, and metabat.2.

-   **Campylobacterota (5 bins)** — Sulfur-oxidizing/cycling bacteria. Present in metabat.56, metabat.32, metabat.43, metabat.14, and metabat.3.

-   **Bacteroidota (3 bins)** — Present in contig_119, metabat.48, and metabat.11.

-   **Firmicutes_A (2 bins)** — Probably heterotrophs. Present in metabat.24 and metabat.47.

-   **Cyanobacteria (1 bin)** — *Roseofilum reptotaenium* at 97.56% of the reference genome, the BBD-dominant cyanobacterium. Only present in metabat.50.

Going by bin count, sulfur-cycling organisms dominate the community, which is consistent with the structure of BBD.

3. **Give a brief summary of the results of prokka and minpath annotation on this assembly. How many genes were identified by Prokka? Of these how many were assigned an EC number?**

```{bash, eval=FALSE, include=FALSE}
cd prokka
ls
ls contig_119
ls contig_119/contig_119.tsv

# while loop reads each line, assigns variable bin to be the value on each line
while read bin;do echo $bin; done < ../checkm/bbd_topbins.txt
# each one is a directory in our prokka directory. we want a pattern line (e.g. name_of_bin, name_of_bin.tsv).

cat ../minpath/*.ec|wc -l
```

```{r prokka EC count, echo=FALSE}
prokka_ECcount <- readxl::read_xlsx("../prokka_minpath_result.xlsx", sheet = 1)
datatable(prokka_ECcount, options = list(scrollX = TRUE, scrollCollapse = TRUE, paging = FALSE), rownames = FALSE,
          caption = "Number of protein-coding genes and genes with an Enzyme Commission (EC) number.")
```

<br/> Across the top 16 bins, 20,257 genes were identified by prokka, of which 5,196 (25%) were assigned an EC number. The rest of the genes are either non-enzymatic (structural, regulatory, or transport proteins) or enzymes without an EC classification. This is a normal proportion for prokaryotic genome annotation. Notably, the proportion of genes with an EC number is fairly stable (\~22–36%) regardless of bin completeness, suggesting Prokka's annotation rate isn't biased by fragmentation — though the absolute number of annotated pathways minpath can detect is still limited by how complete each genome is.

Feeding these EC numbers into minpath produced pathway reports of very different sizes depending on genome completeness:

```{r distinct pathways for top 16 genes, echo=FALSE}
pathways <- readxl::read_xlsx("../prokka_minpath_result.xlsx", sheet = 2)
datatable(pathways, options = list(scrollX = TRUE, scrollCollapse = TRUE, paging = FALSE), rownames = FALSE,
          caption = "Number of unique enzyme coding pathways.")
```

<br/> Pathway richness tracks closely with genome completeness and size — the two \>97%-complete bins (metabat.50, contig_119) have a far richer metabolic annotation. This is a useful caveat for interpreting the lower-completeness bins: absence of a pathway there more likely reflects incomplete genome recovery than a real absence of that metabolic capacity.

## Discussion

**Comment on the quality of your assembly compared with previously published results using Illumina data from Sato et al. 2025. Can you identify bins representing key bacterial taxa that are found in both assemblies? Are there any highly complete bins in one assembly that do not have a clear match in the other assembly? For bins that match across assemblies how to the completeness and contamination statistics compare?**

Sato et al. assembled short-read metagenomes with curated bins using CheckM with tetranucleotide composition and differential cover. If we compare taxonomy at the genus level, there is clear overlap with this nanopore assembly:

-   **Cyanobacteria** - Sato et al.'s dominant BBD cyanobacterium ("Cya2") shared 99% 16S identity with a previously sequenced Caribbean BBC cyanobacterium identified as *Roseofilum reptotaenium*. This directly matches metabat.50, which was independently identified via GTDB-Tk as *Roseofilum reptotaenium* at 97.56% completeness.
-   **Sulfate reducers** - Sato found sulfate-reducing genes in BBD taxonomically dominated by Deltaproteobacteria, typically associated with *Desulfovibrio spp.*, previously believed to be the main sulfide producer in BBD lesions. This matches metabat.27 and metabat.23 (genus *Maridesulfovibrio*, closely related to *Desulfovibrio*).
-   **Sulfide oxidisers/cyclers** - Sato found sulfide-oxidation genes predominantly expressed by Epsilonproteobacteria in BBD (now reclassified as Campylobacterota), with *Arcobacter* as the dominant genus. This assembly recovered several Campylobacterota bins in the family Arcobacteraceae (metabat.56, metabat.32, metabat.43 - genus *Halarcobacter*), a family-level match even though the genus is different.

contig_119 doesn't correspond to any taxon Sato et al. highlighted as a major BBD component - their study actually found that Bacteroidetes were prominent in the precursor "cyanobacterial patch" stage rather than in BBD itself. Recovering such a highly complete Bacteroidota genome directly from a BBD lesion is worth mentioning — it may reflect genuine site-to-site variation (Orpheus Island vs. Pelorus Island, GBR) in BBD community composition, or could simply be that long-read assembly resolved a genome that fragmented out of reach in a short-read assembly.

Completeness and contamination statistics for bins that match Sato et al. were very similar: for Cyanobacteria, my completeness was 97.56% in metabat.50 vs Sato's 98.22% in MAG07, and contamination was 0.67% for both. Qualitatively speaking, recovering two MAGs above 96% completeness with under 2% contamination from a single nanopore sequencing run is a strong outcome. Short-read assemblies often fragment complex or highly repetitive genomes (especially filamentous cyanobacteria like *Roseofilum*), so the near-single-contig, near-complete recovery achieved here for both dominant taxa likely represents a genuine improvement in assembly contiguity over the Illumina-based approach.

**Comment on the results of pathway analysis with minpath. Which bin contains most of the genes required for photosynthesis? Which bin or bins are most likely to be contributing to sulfate reduction?**

Photosynthesis I (PWY-101) was flagged in four bins: contig_119, metabat.27, metabat.48, and metabat.50. Of these, only metabat.50 is a possible carrier of this pathway - the other three have no capacity for oxygenic photosynthesis. Most of the non-cyanobacterial hits reported only 1-2 of the pathway's 4 required gene families found, indicating partial matches from individual genes shared across unrelated pathways (e.g. common electron-transport components) rather than genuine photosynthetic capacity. Therefore, only metabat.50 contains the genes required for photosynthesis, consistent with its taxonomic identity as the BBD-dominant cyanobacterium.

Sulfate/sulfite reduction (PWY-5360, the superpathway of tetrathionate reduction) was found only in metabat.56 - *Halarcobacter* (family Arcobacteraceae) and Campylobacterota, a genus more often associated with sulfide oxidation than reduction. This is an unexpected result; Campylobacterota are known to cycle sulfur compounds depending on conditions, so while this finding isn't unrealistic, it's not a common sulfate reducer.

None of the well-known sulfate-reducing bins identified taxonomically in this assembly - the Desulfobacterota bins metabat.23 and metabat.27 (*Maridesulfovibrio*), and metabat.41, metabat.42, metabat.2 — appear in the PWY-5360 results. This doesn't mean that sulfate-reduction capability was absent as we asked minpath about tetrathionate reduction specifically, and our likely sulfate reducers may simply run a different, unqueried pathway instead. Taxonomically, metabat.23 and metabat.27 are the strongest candidates for driving sulfate reduction in this community, since *Maridesulfovibrio* is closely related to *Desulfovibrio* — the genus Sato et al. identified as the primary sulfide producer in BBD lesions. This suggests minpath's single-pathway search under-represents the true sulfate-reducing potential of the assembly, and a broader search would likely involve more bins.